---
title: "Divvy Data Project"
author: "Matthew Mahoney"
output:
  html_document:
    code_folding: show
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE)
```

## Import libraries

```{r import-libraries, results = 'hide'}
library(tidyverse)
library(remotes)
library(ggmap)
library(lubridate)
library(maps)
library(ggthemes)
library(ggrepel)
library(data.table)
library(broom)
library(gridExtra)
library(knitr)
```

```{r, install stadiamaps, class.source = 'fold-hide'}
# Required if you have not pulled the most recent changes to ggmap
# remotes::install_github('https://github.com/stadiamaps/ggmap')
```

```{r functions-chunk, echo=FALSE}

### create function to find mode

getmode <- function(x){
  u <- unique(x)
  t <- tabulate(match(x, u))
  return(u[t == max(t)])
}

### Donut graphs require same inputs, created function to clean up code

createDonut <- function (x, filter_text){
  
  x %>%
    filter(member_type == filter_text) -> x_prime
  
  x_prime$fraction <- x_prime$ride_count/
    sum(x_prime$ride_count)
  x_prime$ymax <- cumsum(x_prime$fraction)
  x_prime$ymin <- c(0, head(x_prime$ymax, n=-1))
  x_prime$labelPosition <- (x_prime$ymax + x_prime$ymin)/2
  x_prime$season <- c('Winter', 'Spring', 'Summer', 'Fall')
  
  return(x_prime)
  
}
  
getDateString <- function(x){

  daysofweek <- c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday',
                  'Friday', 'Saturday')
  
  return(daysofweek[x])
    
}


### Trip totals are calculated the same for both tables
### created function to reduce redundancy

trip_totals <- function (table_in, filter_name){
  
  ## Create Top_ten_trips table including both counts and lat/lon
  ## coordinates
  
  table_in %>%
    filter(member_type == filter_name) -> a
  
  merge(x = a, y = r_station_pairs,
        by = 'pair_id') %>%
    arrange(desc(trip_count)) %>%
    subset(select = c('start_station_id', 'end_station_id', 'trip_count')) %>%
    head(10) -> b
  
  
  ##Add coordinate information to top_ten_stations using r_station_lat_lng
  merge(x = r_station_lat_lng, y = b,
        by.x = 'station_id', by.y = 'start_station_id') %>%
    merge(y = r_station_lat_lng, by.x = 'end_station_id', 
          by.y = 'station_id') -> b
  
  b <- setcolorder(b, c(2,3,4,1,6,7,5))
  
  return(b)
}
```

```{R theme-preset, echo=FALSE}
## create theme preset for future graphs
baseTheme <-   theme_classic() +
  theme(plot.title = element_text(hjust = .5, size = 13, face = 'bold',
                                  family = 'serif'),
        panel.background = element_rect(fill = "#3f403f"),
        plot.background =  element_rect(fill = "#BEBEBE"),
        axis.text.x = element_text(color =  "black", family = 'serif'),
        axis.text.y = element_text(color =  "black", family = 'serif'),
        strip.text = element_text(size = 11, family = 'serif'))

axisScale <- 1000
yScaleMax <- 1000
yScaleMin <- 0
byValue <- 50


## x-axis values for graphs
discreteX <- scale_x_discrete(name = 'day of week',
                 limits = c('sun','mon','tue','wed','thur','fri','sat'))
## y-axis values for graphs
continuousY <-  scale_y_continuous(name = paste("total_rides (",
                                                axisScale, '\'s)'),
                     seq(yScaleMin, yScaleMax,by=byValue),
                     labels = scales::comma)

```

## Data Import

```{r data-import, class.source='fold-hide'}

## Data Import

list.files(pattern = 'd_ride') %>%
  map_df(~fread(header = FALSE, .)) -> d_ride_table

r_station_lat_lng <- read.csv(
"cyclistic_r_station_lat_lon.csv",
                              header = FALSE)
r_station_table <- read.csv(
  "cyclistic_r_station_table.csv", 
                            header = FALSE)

setDT(d_ride_table)

### rename columns

names(r_station_lat_lng) <- c('station_id', 'lat', 'lon')
names(d_ride_table) <- c('ride_id', 'rideable_type', 'started_at', 'ended_at',
                         'start_station_id', 'end_station_id',
                         'member_casual', 'ride_in_seconds',
                         'ride_length', 'year_of_ride', 'month_of_ride',
                         'day_of_week')
names(r_station_table) <- c('station_name', 'station_key')

### reclassify datetimes as POSIX
d_ride_table$started_at <- as_datetime(d_ride_table$started_at)
d_ride_table$ended_at <- as_datetime(d_ride_table$ended_at)
```

```{R table-creation, include=FALSE}

###Table Creation

## create ride table with only casual member entries    
d_ride_table %>% filter(member_casual == 'casual') -> d_ride_table_c

setDT(d_ride_table_c)


## create ride table with only member entries
d_ride_table %>% filter(member_casual == 'member') -> d_ride_table_m

setDT(d_ride_table_m)


#create a table containing all start/end station pairs in the data set
d_ride_table %>%
  subset(select = c('start_station_id', 'end_station_id')) %>%
  unique() -> r_station_pairs


#create unique id for each pair of start/end stations
r_station_pairs$pair_id <- c(1:nrow(r_station_pairs)) 
r_station_pairs <- setcolorder(r_station_pairs, c(3,1,2)) #reorder columns


## Create final station pair table with station_names
merge(x = r_station_pairs, y = r_station_table,
      by.x = 'start_station_id', by.y = 'station_key') %>%
  merge(y = r_station_table, by.x = 'end_station_id', by.y = 'station_key') -> 
  r_station_pairs
r_station_pairs <- setcolorder(r_station_pairs,c(3,4,2,5,1)) #reorder columns
names(r_station_pairs)[2] <- 'start_station_name' #rename column
names(r_station_pairs)[4] <- 'end_station_name' #rename column

setDT(r_station_pairs)


#create table containing all rides and include start/end station pair ids
d_ride_table %>%
  subset(select = c('ride_id', 'start_station_id', 'end_station_id', 
                    'member_casual')) -> d_ride_table_prime

merge(x=d_ride_table_prime, y=r_station_pairs,
      by.x = c('start_station_id', 'end_station_id'),
      by.y = c('start_station_id', 'end_station_id')) -> d_station_pair_merge



d_station_pair_merge <- setcolorder(d_station_pair_merge,
                                    c(3,4,1,2)) #reorder columns
rm(d_ride_table_prime)


#create table with total trips taken from all station pairs
  d_station_pair_merge[, .(.N), by = .(pair_id, 
                       member_casual)] -> d_station_trip_totals
  
names(d_station_trip_totals) <- c('pair_id', 'member_type', 'trip_count')
setcolorder(d_station_trip_totals, c(1,3,2)) #reorder columns


#Create table that counts bike type by member status.
#Removed docked_bike option as it is not used by members
#So there is no comparison to be made
rideable_type_count <- d_ride_table %>%
  subset(select = c(rideable_type, member_casual)) %>%
  group_by(rideable_type, member_casual) %>%
  summarize(Count = length(rideable_type)) %>%
  filter(rideable_type != "docked_bike")

```

## Analysis{.tabset}

### Plots

Following data import and table creation, I begin my analysis finding
the average ride length and the mode of the day of week for 
each group using the service.

```{r, results=FALSE}

## find avg length of ride and the day of week mode for casual riders
mean(d_ride_table_c$ride_in_seconds)/60 -> avg_ride_length_casual 
getmode(d_ride_table_c$day_of_week) -> mode_dow_c


## find avg length of ride and the day of week mode for members
mean(d_ride_table_m$ride_in_seconds)/60 -> avg_ride_length_member
getmode(d_ride_table_m$day_of_week) -> mode_dow_m

mode_dow_c <- getDateString(mode_dow_c)
mode_dow_m <- getDateString(mode_dow_m)

```

We find that for casual riders, the most common riding day is
`r mode_dow_c`, and the most common day for members is `r mode_dow_m`.

This is an interesting insight already, and leads me to believe that it
may be possible that casual riders are riding more on weekends because
they do not use the service to commute to a job.

Continuing on this path, I create a table that contains the average ride
length by day for both members and casual riders, and also one that
checks the number of rides by day of week over 12 months of data.

```{r, include=FALSE}
# create data frame for day and ride length information
dow <- c(1,1,2,2,3,3,4,4,5,5,6,6,7,7)
d_avg_ride_by_day <- data.frame(matrix(nrow = length(dow), ncol = 3))
colnames(d_avg_ride_by_day) = c('day', 'ride_length', 'member_type')

d_avg_ride_by_day$day <- dow
```

```{r}

## find avg ride length by day for members and store in a table
d_ride_table %>%
  subset(select = c(day_of_week, ride_in_seconds, member_casual)) %>%
  group_by(day_of_week, member_casual) %>%
  summarize(avg_ride = round(mean(ride_in_seconds)/60,
                             digits= 3)) -> a

d_avg_ride_by_day$ride_length <- a$avg_ride 
d_avg_ride_by_day$member_type <- a$member_casual 

rm(a)
```

Graphing these leads to an interesting insight.

```{R echo=FALSE}

## Plot Average ride length for members and casual riders
d_avg_ride_by_day %>%
  ggplot(aes(x = day, y = ride_length)) +
  ggtitle("Avg Ride Length, Member vs Casual Monday to Sunday") +
  geom_col(position = position_dodge(.9), aes(fill = member_type),
           color = 'grey') +
  discreteX +
  scale_y_continuous(name = "Ride_Length (mins)",
                     seq(0,30,by=5), labels = scales::number) +
  geom_text(aes(label = round(ride_length,2), group = member_type),  
            color = 'black', size = 3, position = position_dodge(0.9),
            vjust = 1.5) +
  scale_fill_brewer(palette = "Paired") +
  baseTheme

```

```{R, echo=FALSE}

## Plot member ride count for the last 12 months by weekday
ggplot(d_ride_table,
       aes(x=day_of_week)) +
  ggtitle("Rides by Weekday Monday to Sunday by Membership Type") +
  geom_bar(
    aes(y = after_stat(count)/axisScale),
    fill = '#A1CDF4',
    color = 'black',
    alpha = 0.8) +
  discreteX +
  continuousY +
  geom_text(
    aes(y = after_stat(count)/axisScale, 
        label = round(after_stat(count)/axisScale,1)),
    stat="count", vjust = 1.5, size = 3) +
  facet_wrap(~member_casual) +
  baseTheme

```

The above graphs show that members take shorter more consistent rides,
and also ride more frequently during the week days, while casual riders
take longer more varied rides, and ride more during the weekends. This
makes a strong case that members use the service for activities like
commuting or scheduled exercise, while casual riders likely use the
service more commonly for joy riding, or weekend activities.

Next I want to take a look at seasonal ridership, to do this, I create a
new table that contains a count of rides taken by the two rider types in
each season.

```{r, echo=FALSE}
sm <- -2 #start month
em <- 0 #end month

## loop through filtered tables to find the number of 
## rides in different seasons. increment sm & em by 3 to create
## seasonal ranges
## add the results to the rides_by_season table

#generate empty vectors to fill the columns in rides_by_season table 
season <- c()
ride_count <- c()
member_type <- c()
```
Additional code hidden here.
```{r, results=FALSE, class.source = 'fold-hide'}

## Loop to generate values for season, ride_count, and member_type
for (j in 1:4){ #begin for
   
  sm <- sm + 3 ## add 3 to both values to get next season range
  em <- em + 3
  

d_ride_table_c %>%
  filter(month_of_ride >= sm & month_of_ride <= em) %>% ## filter for month range
  summarize(count_r = n()) -> a

ride_count <- append(ride_count, a$count_r) ##append to vectors
season <- append(season, j)
member_type <- append(member_type, 'casual')

d_ride_table_m %>%  
  ## filter for month range
  filter(month_of_ride >= sm & month_of_ride <= em) %>% 
  summarize(count_r = n()) -> b

ride_count <- append(ride_count, b$count_r) ##append to vectors
season <- append(season, j)
member_type <- append(member_type, 'member')
  
  rm(a,b)
  
} # end for

```

```{r}

#create new data frame with vectors
d_rides_by_season <- data.frame(season, ride_count, member_type) 

rm(sm,em,j, season, ride_count, member_type)
```

```{r, include=FALSE}
gc()

```

Visualizing this data, we find what we would likely expect, especially
considering the weather patterns in a place like Chicago.

```{R, echo=FALSE}

##create tables for donut graphs
d_rides_by_season_c <- createDonut(d_rides_by_season, 'casual')
d_rides_by_season_m <- createDonut(d_rides_by_season, 'member')


##create donut graph visual preset
donutTheme <-  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5,size = 20, face = 'bold',
                                  family = 'serif'))+
  theme(plot.background = element_rect(fill = "#f0f0f0") ) +
  theme(legend.position = 'none')


## plot donut graphs for casual riders and members
## for rides per season
ggplot(d_rides_by_season_c, aes(ymax = ymax, ymin = ymin, xmax = 4,
                                xmin =3, fill = season)) +
  ggtitle("Rides by Season Casual") +
  geom_rect()+
  coord_polar(theta = 'y') +
  xlim(c(0.5,4))+
  scale_fill_brewer(palette = "Spectral") +
  geom_text(x=3.5, aes(y = labelPosition, label = season, fontface = 'bold'), size=3) +
  geom_text(x = 4.5, aes(y = labelPosition-0.02,
                         label = paste(round(fraction*100, 2), '%')), size=2.5) +
  
  donutTheme -> g_1 ## assign graph 1
  
ggplot(d_rides_by_season_m, aes(ymax = ymax, ymin = ymin, xmax = 4,
                                xmin =3, fill = season)) +
  ggtitle("Rides by Season Member") +
  geom_rect()+
  coord_polar(theta = 'y') +
  xlim(c(0.5,4))+
  scale_fill_brewer(palette = "Spectral") +
  geom_text(x=3.5, aes(y = labelPosition, label = season, fontface = 'bold'), size=3) +
  geom_text(x = 4.5, aes(y = labelPosition+0.02,
                         label = paste(round(fraction*100, 2), '%')),size = 2.5) +
  
  donutTheme -> g_2 ## assign graph 2

grid.arrange(g_1, g_2, ncol=2, respect = TRUE)
```

We find that generally ridership is dispersed equally for casual riders
and members. However, members do ride more in colder months than casual
riders do. Though this is a relatively small difference, it could
suggest that members do choose to ride more even in the colder months as
they use the service for necessary tasks like commuting.

The final relationship I'd like to take a look at is the types of bikes
that casual riders and members use.

```{R, echo=FALSE}

##Plot bike type usage for both casual and member riders
rideable_type_count %>%
  ggplot(aes(rideable_type, Count/1000)) +
  ggtitle("Bike Type Usage Member vs Casual") +
  geom_col(position = position_dodge(.8),
           width = .75,
           aes(fill = member_casual), 
           color = 'grey') +
  geom_text(aes(label = round(rideable_type_count$Count/1000, 1), 
                group = member_casual),
            position = position_dodge(.8), vjust = 1.5) +
  xlab('Bike Type') +
  ylab("Ride count (1000's)") +
  scale_fill_brewer(palette = "Paired") +
  baseTheme

```

```{r, echo=FALSE}
rideable_type_count %>%
  group_by(member_casual) %>%
  mutate(percent = round(Count/sum(Count),4)*100) -> a
```

Casual members use classic bikes `r a$percent[1]`% of the time and
electric bikes `r a$percent[3]`% of the time, while members use classic
bikes `r a$percent[2]`% of the time and electric bikes `r a$percent[4]`%
of the time.

This is also reflected in the visualization of the counts above. We can
clearly see that even though members use the service at higher rates,
their usage shows a preference to classic bikes.

Another thing to look at is rides during time of day. I start by taking
a large sample of 100,000 from the main dataset d_ride_table and create 
a plot for which hours of the day casual riders/members use bikes the most.

```{r, class.source='fold-hide'}
d_ride_table_sample <- sample_n(d_ride_table,100000) %>%
  subset(select = c(member_casual, started_at,ended_at)) %>%
  mutate(hour_of_ride = format(as.POSIXct(started_at), format = "%H"))

d_ride_table_sample$hour_of_ride <- as.numeric(d_ride_table_sample$hour_of_ride)

d_hour_of_day_table <- d_ride_table_sample %>%
  group_by(member_casual) %>%
  count(hour_of_ride) %>%
  mutate(ridesPercent = n/sum(n)*100)
```

```{r}
##Plot bike type usage for both casual and member riders by percent
d_hour_of_day_table %>%
    ggplot(aes(x=hour_of_ride, y = ridesPercent, fill = member_casual,
                                  group = member_casual)) +
    geom_bar(stat = 'identity', position = position_dodge(0.9)) +
    scale_x_continuous(breaks = seq(0,23,1)) +
    ylim(0,12) +
    scale_fill_brewer(palette = "Pastel1") +
    labs(x = 'Departure Time', y = 'Percent of Rides', 
         fill = 'Membership Status' ) +
    theme_bw()

```

### Mapping


```{r, class.source = 'fold-hide', results = FALSE}
## Create tables for the top ten trips for casual and member riders

top_ten_trips_c <- trip_totals(d_station_trip_totals, 'casual')
top_ten_trips_m <- trip_totals(d_station_trip_totals, 'member')
```

```{r, class.source = 'fold-hide', results = FALSE}

## using openstreetmap for bounding box info for

chicago_map <- get_stadiamap(
  bbox = c(left = -87.73, bottom = 41.76, right = -87.57, top = 41.97),
  maptype='stamen_terrain',
  zoom = 13) %>% ggmap()
```

```{r}

## Map all stations within the bounds of the top 10         
chicago_map +
  geom_point(data = r_station_lat_lng, aes(x = lon, y = lat))+
  ggtitle(label = 'Station Locations')
```

```{r}

## Map start/end station locations for the top 10 trips taken by members
chicago_map +
  geom_point(data = top_ten_trips_m
             ,aes(x = lon.x, y=lat.x),size = 2)+
  ggtitle(label = 'Top 10 Most Visited Stations Members') +
  geom_point(data = top_ten_trips_m,
             aes(x = lon.y, y = lat.y),size =2)+
  theme(legend.position = "none")
```

```{r}

## Map start/end station locations for the top 10 trips taken by
## casual riders
chicago_map +
  geom_point(data = top_ten_trips_c
             ,aes(x = lon.x, y=lat.x),size = 2)+
  ggtitle(label = 'Top 10 Most Visited Stations Casual Riders') +
  geom_point(data = top_ten_trips_c,
             aes(x = lon.y, y = lat.y),size =2)+
  theme(legend.position = "none")
```

```{r, include=FALSE}
gc()
```

### Table Index

```{r, echo=FALSE}

## ride table with only casual member entries    

knitr::kable(head(d_ride_table_c,5))


knitr::kable(head(d_ride_table_m,5))

knitr::kable(head(r_station_pairs,5))

knitr::kable(head(d_station_trip_totals,5))

knitr::kable(head(d_rides_by_season,5))

knitr::kable(head(d_avg_ride_by_day))
```

```{r, eval=FALSE}
## create ride table with only member entries
d_ride_table %>% filter(member_casual == 'member') -> d_ride_table_m

setDT(d_ride_table_m)


#create a table containing all start/end station pairs in the data set
d_ride_table %>%
  subset(select = c('start_station_id', 'end_station_id')) %>%
  unique() -> r_station_pairs


#create unique id for each pair of start/end stations
r_station_pairs$pair_id <- c(1:nrow(r_station_pairs)) 
r_station_pairs <- setcolorder(r_station_pairs, c(3,1,2)) #reorder columns


## Create final station pair table with station_names
merge(x = r_station_pairs, y = r_station_table,
      by.x = 'start_station_id', by.y = 'station_key') %>%
  merge(y = r_station_table, by.x = 'end_station_id', by.y = 'station_key') -> 
  r_station_pairs
r_station_pairs <- setcolorder(r_station_pairs,c(3,4,2,5,1)) #reorder columns
names(r_station_pairs)[2] <- 'start_station_name' #rename column
names(r_station_pairs)[4] <- 'end_station_name' #rename column

setDT(r_station_pairs)


#create table containing all rides and include start/end station pair ids
d_ride_table %>%
  subset(select = c('ride_id', 'start_station_id', 'end_station_id', 
                    'member_casual')) -> d_ride_table_prime

merge(x=d_ride_table_prime, y=r_station_pairs,
      by.x = c('start_station_id', 'end_station_id'),
      by.y = c('start_station_id', 'end_station_id')) -> d_station_pair_merge



d_station_pair_merge <- setcolorder(d_station_pair_merge,
                                    c(3,4,1,2)) #reorder columns
rm(d_ride_table_prime)


#create table with total trips taken from all station pairs
  d_station_pair_merge[, .(.N), by = .(pair_id, 
                       member_casual)] -> d_station_trip_totals
  
names(d_station_trip_totals) <- c('pair_id', 'member_type', 'trip_count')
setcolorder(d_station_trip_totals, c(1,3,2)) #reorder columns


#Create table that counts bike type by member status.
#Removed docked_bike option as it is not used by members
#So there is no comparison to be made
rideable_type_count <- d_ride_table %>%
  subset(select = c(rideable_type, member_casual)) %>%
  group_by(rideable_type, member_casual) %>%
  summarize(Count = length(rideable_type)) %>%
  filter(rideable_type != "docked_bike")

```

### Functions Index

```{r, eval=FALSE}

### create function to find mode

getmode <- function(x){
  u <- unique(x)
  t <- tabulate(match(x, u))
  return(u[t == max(t)])
}

### Donut graphs require same inputs, created function to clean up code

createDonut <- function (x, filter_text){
  
  x %>%
    filter(member_type == filter_text) -> x_prime
  
  x_prime$fraction <- x_prime$ride_count/
    sum(x_prime$ride_count)
  x_prime$ymax <- cumsum(x_prime$fraction)
  x_prime$ymin <- c(0, head(x_prime$ymax, n=-1))
  x_prime$labelPosition <- (x_prime$ymax + x_prime$ymin)/2
  x_prime$season <- c('Winter', 'Spring', 'Summer', 'Fall')
  
  return(x_prime)
  
}
  
getDateString <- function(x){

  daysofweek <- c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday',
                  'Friday', 'Saturday')
  
  return(daysofweek[x])
    
}


### Trip totals are calculated the same for both tables
### created function to reduce redundancy

trip_totals <- function (table_in, filter_name){
  
  ## Create Top_ten_trips table including both counts and lat/lon
  ## coordinates
  
  table_in %>%
    filter(member_type == filter_name) -> a
  
  merge(x = a, y = r_station_pairs,
        by = 'pair_id') %>%
    arrange(desc(trip_count)) %>%
    subset(select = c('start_station_id', 'end_station_id', 'trip_count')) %>%
    head(10) -> b
  
  
  ##Add coordinate information to top_ten_stations using r_station_lat_lng
  merge(x = r_station_lat_lng, y = b,
        by.x = 'station_id', by.y = 'start_station_id') %>%
    merge(y = r_station_lat_lng, by.x = 'end_station_id', 
          by.y = 'station_id') -> b
  
  b <- setcolorder(b, c(2,3,4,1,6,7,5))
  
  return(b)
}
```
